---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# biocthis

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/biocthis.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/biocthis)
[![BioC dev status](http://www.bioconductor.org/shields/build/devel/bioc/biocthis.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/biocthis)
[![R build status](https://github.com/lcolladotor/biocthis/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/lcolladotor/biocthis/actions)
[![Codecov test coverage](https://codecov.io/gh/lcolladotor/biocthis/branch/master/graph/badge.svg)](https://codecov.io/gh/lcolladotor/biocthis?branch=master)
[![Support site activity, last 6 months: tagged questions/avg. answers per question/avg. comments per question/accepted answers, or 0 if no tagged posts.](http://www.bioconductor.org/shields/posts/biocthis.svg)](https://support.bioconductor.org/t/biocthis/)
[![GitHub issues](https://img.shields.io/github/issues/lcolladotor/biocthis)](https://github.com/lcolladotor/biocthis/issues)
<!-- badges: end -->

`r BiocStyle::Githubpkg("lcolladotor/biocthis")` is an R package that expands `r BiocStyle::CRANpkg("usethis")` with Bioconductor-friendly templates. These templates will help you quickly create an R package that either has Bioconductor dependencies or that you are thinking of submitting to Bioconductor one day. `r BiocStyle::Githubpkg("lcolladotor/biocthis")`  has functions that can also enhance your current R packages that either are already distributed by Bioconductor or have Bioconductor dependencies. `r BiocStyle::Githubpkg("lcolladotor/biocthis")` also includes a Bioconductor-friendly [GitHub Actions](https://github.com/features/actions) workflow for your R package(s). To use the functions in this package, you need to load it.

Note that `r BiocStyle::Githubpkg("lcolladotor/biocthis")` is not a Bioconductor-core package and as such it is not a Bioconductor official package. It was made by and for Leonardo Collado-Torres so he could more easily maintain and create Bioconductor packages as listed at [lcolladotor.github.io/pkgs/](https://lcolladotor.github.io/pkgs/). Hopefully `r BiocStyle::Githubpkg("lcolladotor/biocthis")`  will be helpful for you too.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `biocthis` using from [Bioconductor](http://bioconductor.org/) the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("biocthis")
```

And the development version from [GitHub](https://github.com/) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("lcolladotor/biocthis")
```

## Example

Here is how you can use `r BiocStyle::Githubpkg("lcolladotor/biocthis")` to create a new Bioconductor-friendly R package (illustrated using a temporary directory):

```{r example, eval = requireNamespace('biocthis')}
## Load biocthis
library("biocthis")

## Set the package name
pkgname <- "biocthisexample"

## Create the example package in a temporary location
withr::with_dir(tempdir(), {
    usethis::create_package(pkgname)
})

## Save the path to our temporary package for the rest of the examples
pkgdir <- file.path(tempdir(), pkgname)

## Create the bioc templates
withr::with_dir(pkgdir, {
    biocthis::use_bioc_pkg_templates()
})
```

The template `dev` scripts include comments and steps you can follow for making your Bioconductor-friendly R package or updating a current package. In particular, you might
want to use a Bioconductor-friendly GitHub Actions workflow. If this is your first time seeings this words, we highly recommend that you watch [Jim Hester's `rstudio::conf(2020)` talk on this subject](https://www.jimhester.com/talk/2020-rsc-github-actions/).

```{r 'bioc_github_action'}
## Create a GitHub Actions workflow that is Bioconductor-friendly
withr::with_dir(pkgdir, {
    biocthis::use_bioc_github_action()
})
```

Alternatively, use `usethis::use_github_action()` for the general GitHub Actions workflow maintained by `r-lib/actions` or `usethis::use_github_action("check-bioc", "https://bit.ly/biocthis_gha")` without having to install `r BiocStyle::Githubpkg("lcolladotor/biocthis")`.


## Citation

Below is the citation output from using `citation('biocthis')` in R. Please
run this yourself to check for any updates on how to cite __biocthis__.

```{r 'citation', eval = requireNamespace('biocthis')}
print(citation("biocthis"), bibtex = TRUE)
```

Please note that the `biocthis` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `biocthis` project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, `r BiocStyle::Githubpkg('r-hub/sysreqs')` and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://.github.io/lcolladotor/biocthis) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Githubpkg('lcolladotor/biocthis')`.


