---
title: "{{{ vignette_title }}}"
author: 
  - name: Your name
    affiliation:
    - Your affiliation
    email: your_email@somewhere.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver("{{ Package }}")`"
vignette: >
  %\VignetteIndexEntry{{{ braced_vignette_title }}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("knitcitations")

## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = "to.doc", citation_format = "text", style = "html")
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle"),
    knitcitations = citation("knitcitations"),
    knitr = citation("knitr")[3],
    rmarkdown = citation("rmarkdown"),
    sessioninfo = citation("sessioninfo"),
    testthat = citation("testthat"),
    {{ Package }} = citation("{{ Package }}")[1]
)

write.bibtex(bib, file = "{{ Package }}.bib")
```

# Basics

## Install `{{ Package }}`

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. `r Biocpkg('{{ Package }}')` is a `R` package available via the [Bioconductor](http://bioconductor/packages/{{ Package }}) repository for packages. `R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install `r Biocpkg('{{ Package }}')` by using the following commands in your `R` session:

```{r 'installDer', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
  }

BiocManager::install("{{ Package }}")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Required knowledge

`r Biocpkg('{{ Package }}')` is based on many other packages and in particular in those that have implemented the infrastructure needed for dealing with RNA-seq data (EDIT!). That is, packages like `r Biocpkg('SummarizedExperiment')` (EDIT!).

If you are asking yourself the question "Where do I start using Bioconductor?" you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `{{ Package }}` tag and check [the older posts](https://support.bioconductor.org/t/{{ Package }}/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

## Citing `r Biocpkg('{{ Package }}')`

We hope that `r Biocpkg('{{ Package }}')` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r 'citation'}
## Citation info
citation("{{ Package }}")
```

# Quick start to using to `{{ Package }}`

```{r 'start', message=FALSE}
library("{{ Package }}")
```

Edit this as you see fit =)

Here is an example of you can cite your package inside the vignette:

* `{{ Package }}` `r citep(bib[['{{ Package }}']])`



# Reproducibility

The `r Biocpkg('{{ Package }}')` package `r citep(bib[['{{ Package }}']])` was made possible thanks to:

* R `r citep(bib[['R']])`
* `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
* `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`
* `r CRANpkg('knitr')` `r citep(bib[['knitr']])`
* `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])`
* `r CRANpkg('sessioninfo')` `r citep(bib[['sessioninfo']])`
* `r CRANpkg('testthat')` `r citep(bib[['testthat']])`

This package was developed using `r BiocStyle::Githubpkg('lcolladotor/biocthis')`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("{{ vignette_name }}.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("{{ vignette_name }}.Rmd", tangle = TRUE)
```

```{r createVignette2}
## Clean up
file.remove("{{ Package }}.bib")
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
with `r CRANpkg('knitr')` `r citep(bib[['knitr']])` and `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])` running behind the scenes.

Citations made with `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`.

```{r vignetteBiblio, results = 'asis', echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
